---
title: "SimHashReport"
execute:
  echo: false
format: html
editor: visual
---

# SimHash: Boosting Efficiency in Similarity Search with Minimal Accuracy Loss

Amiya Harish

## Abstract

My research question asks if the loss in accuracy of a similarity search algorithm using SimHash worth the efficiency. I determine this by comparing the performance metrics of an algorithm using SimHash and the basic cosine similarity algorithm. I created a vector for each article in the New York Times articles dataset with the number of times each word in the full word list appears in each article. I then used this vector list to find similarity scores using different algorithms. The dataset is very large, with 300,000 articles, however, in order for my computer to be able to handle the program I took a subset of the first 1000 articles to run the algorithms on. My main finding from this project is that the loss in accuracy with using a SimHash algorithm to calculate similarity between articles is insignificant when compared to the large boost in efficiency. This project proves that in the future, SimHash can transform how we handle large datasets by balancing speed with acceptable accuracy. As a result, further advancements in SimHash could drive even greater efficiency and applicability across new domains, drastically improving what we can do with technology.

## Background

The dataset I use consists of articles from the New York Times and all the words that each article contains. My research question aims to decide whether the loss in accuracy of a similarity search algorithm using SimHash is worth the efficiency. I chose this research question because SimHash is widely recognized as the fastest similarity search algorithm that uses cosine similarity. I wanted to compare algorithms using cosine similarity and not Jaccard similarity because while cleaning my data I found that many articles in the dataset had near duplicate articles and cosine similarity algorithms automatically combine near duplicate articles. This research question is important because as technology improves, we need to work to find the most efficient methods of computing everything- however, we must ensure that this does not come at the cost of too much of loss in accuracy. Before beginning this project, I hypothesized that the loss of accuracy would be worth the efficiency when using SimHash based on my previous research.

## Figures

```{r setup}
#| echo: FALSE
#| message: FALSE

library(caret)
library(scales)
library(digest)
library(stringi)
library(htmlwidgets)
library(lsa)
library(Matrix)
library(textreuse)
library(profvis)
library(data.table) # this library contains the fread() function
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(pheatmap) # for clustering and reordering
```

I first calculated the time and memory performance metrics for each of the algorithms.

```{r}
#| echo: FALSE

# Create the data frame
plot_time_data <- data.frame(
  Algorithm = c("Basic Cosine Similarity", "SimHash"),
  Time = c(15790, 630),
  Memory = c(103.2, 5.5)
)

# Plot for Time
ggplot(plot_time_data, aes(x = Algorithm, y = Time, fill = Algorithm)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Basic Cosine Similarity" = "blue", "SimHash" = "red")) +
  labs(title = "Time Usage by Algorithm",
       x = "Algorithm",
       y = "Time (ms)",
       fill = "Algorithm") +
  theme_minimal()
```

Above is a bar graph that shows the time it takes each algorithm to calculate similarity scores in milliseconds. It is clear that the basic cosine similarity algorithm takes almost 20 times longer than the SimHash algorithm.

```{r}
#| echo: FALSE

# Create the data frame
plot_time_data <- data.frame(
  Algorithm = c("Basic Cosine Similarity", "SimHash"),
  Time = c(15790, 630),
  Memory = c(103.2, 5.5)
)

# Plot for Memory
ggplot(plot_time_data, aes(x = Algorithm, y = Memory, fill = Algorithm)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Basic Cosine Similarity" = "blue", "SimHash" = "red")) +
  labs(title = "Memory Usage by Algorithm",
       x = "Algorithm",
       y = "Memory (bytes)",
       fill = "Algorithm") +
  theme_minimal()
```

Above is a bar graph that shows the memory that each algorithm takes up to calculate similarity scores in bytes. Similarly to time, is clear that the basic cosine similarity algorithm takes up almost 20 times more space than the SimHash algorithm.

Next, I calculated the accuracy of each algorithm.

```{r}
#| echo: FALSE

accuracy_cosine <- 1
accuracy_simHash <- 0.99889

# Create the data frame
plot_data <- data.frame(
  Algorithm = factor(c("Basic Cosine Similarity", "SimHash")),
  Accuracy = c(accuracy_cosine, accuracy_simHash)
)

# Create the lollipop chart
ggplot(plot_data, aes(x = Algorithm, y = Accuracy, color = Algorithm)) +
  geom_segment(aes(x = Algorithm, xend = Algorithm, y = 0.9, yend = Accuracy), size = 1) +
  geom_point(size = 4) +
  scale_y_continuous(limits = c(0.9, 1), breaks = seq(0.9, 1, 0.01), labels = percent) +
  scale_color_manual(values = c("Basic Cosine Similarity" = "blue", "SimHash" = "red")) +
  labs(title = "Accuracy Comparison of Similarity Search Algorithms",
       x = "Algorithm",
       y = "Accuracy") +
  theme_minimal() +
  theme(legend.position = "none")
```

Above is a lollipop graph that shows the accuracy of each algorithm in calculating similarity scores by percentage. The graph is scaled to zoom in on the top 10 percent accuracy. It is clear that the accuracy of SimHash is only slightly lower in comparison with the basic cosine similarity algorithm which has 100 percent accuracy.

## Discussion

This proejct demonstrates the potential of the SimHash algorithm for optimizing similarity searches in large datasets. The comparison between SimHash and the basic cosine similarity algorithm reveals a striking trade-off between efficiency and accuracy. While the basic cosine similarity algorithm achieves perfect accuracy, the SimHash algorithm, with a slightly lower accuracy of 99.89 percent, offers a substantial boost in both time and memory efficiency. This balance suggests that in many practical applications, the minor loss in accuracy is justifiable given the significant gains in performance.

SimHash proves to be a highly efficient alternative for similarity searches, particularly in scenarios requiring rapid processing or limited computational resources. The ability to maintain near-perfect accuracy while significantly reducing time and memory usage makes SimHash particularly relevant in data science, where large-scale data processing is common.

This research highlights the growing importance of balancing accuracy with efficiency in data science. As datasets continue to expand, methods like SimHash that optimize resource usage without compromising too much on accuracy will become increasingly important.

A key limitation of this project is the use of a small subset of the New York Times dataset, which may not fully represent SimHash's performance across more diverse and larger datasets. Additionally, while the accuracy loss is minimal, it may not be acceptable in applications where precision is critical, such as medical or legal domains.

Future research should explore SimHash’s performance with larger and more varied datasets to validate these findings. Could combining SimHash with other algorithms further enhance efficiency without sacrificing accuracy?

This project proves the necessity for the continued use and development of SimHash in data science, particularly where efficiency is prioritized and slight accuracy trade-offs are acceptable.

## **Code and Data Availability**

All my scripts, code, and plots can be accessed through this link to my GitHub repository: <https://github.com/the-codingschool/DSRP-2024-Alex/tree/570787a68e34cc6128ea614e8098311f1b6fa5bd/DRSP-2024-Amiya>

The dataset that I worked with came from the UC Irvine Machine Learning Repository Archive: <https://archive.ics.uci.edu/dataset/164/bag+of+words>

## Acknowledgements

I would like to acknowledge Professor Alex Andoni for introducing the dataset, providing ideas, feedback, and helping me throughout this project.

I would like to acknowledge Sarah Parker for teaching me the basics of Data Science, coding in R, and how to write a research paper.

I would like to acknowledge Pavithra, the TA that I worked with, for teaching and helping me solve any issues that I encountered while coding.

I would like to acknowledge Alex's research group for sharing ideas and discussing issues throughout the process.

I would like to acknowledge the DSRP team, the Coding School, and Columbia University for holding this program and giving me this opportunity, providing me with all the resources I needed to complete this project.
